---
title: "Reproducible Research: Peer Assessment 1"
output: 
html_document:
keep_md: true
---
## By Joanna Widjaja, 15 Feb 2015 

## Loading and preprocessing the data
```{r}
unzip("activity.zip")
activity <- read.csv("activity.csv")
```

## What is mean total number of steps taken per day?
```{r}
aggActivityDate <- aggregate(steps ~ date, data = activity, FUN = sum)
hist(aggActivityDate$steps, main = "Total Number of Steps", xlab = "Total Number of Steps by Day")
```

Below is the mean and median of the total number of steps taken per day:
```{r}
mean(aggActivityDate$steps)
median(aggActivityDate$steps)
```

## What is the average daily activity pattern?
```{r}
aggActivityInt <- aggregate(steps ~ interval, data = activity, FUN = mean, na.rm = TRUE)
library(ggplot2)
ggplot(aggActivityInt, aes(interval, steps)) + geom_line(colour="#BB0000") + labs(title="Mean of Steps by Interval")
```

Below is the interval that has the maximum mean number of steps:
```{r}
aggActivityInt[aggActivityInt$steps == max(aggActivityInt$steps), 1]
```

## Inputing missing values
Below is the number of missing rows:
```{r}
sum(is.na(activity$steps))
```

Below is the new dataset with missing values replaced:
```{r}
library(plyr)
newActivity <- join(activity, aggActivityInt, by = 'interval')
newActivity[is.na(newActivity$steps),1] <- newActivity[is.na(newActivity$steps),4]
newActivity <- newActivity[,c(1,2,3)]
head(newActivity)
```

Graph new histogram.
```{r}
aggNewActivity <- aggregate(steps ~ date, data = newActivity, FUN = sum)
hist(aggNewActivity$steps, main = "Total Number of Steps", xlab = "Total Number of Steps by Day")
```

Calculate mean and median.
```{r}
mean(aggNewActivity$steps)
median(aggNewActivity$steps)
```

## Are there differences in activity patterns between weekdays and weekends?
```{r}
newActivity$wday <- ifelse(0 < as.POSIXlt(newActivity$date)$wday & as.POSIXlt(newActivity$date)$wday < 6, 1, 0)
newActivity$wday <- factor(newActivity$wday, levels = c(0,1), labels = c("weekend", "weekday"))
aggNewActivity <- aggregate(steps ~ interval + wday, data = newActivity, FUN = mean)
library(lattice)
xyplot(steps ~ interval | factor(wday), data = aggNewActivity, aspect = 1/2,type = "l")
```

